Sending a Balloon to the Edge of Space

### 2018-11-13 paper/paper.tex
start knitting paper

---
title: "Sending a Balloon to the Edge of Space"
author: Luker Bowsher, John Kim, Vivian Liu, Simon Oros, Caroline Pang, Alec Vercruysse
date: "November 13, 2018"
output: 
  pdf_document:
    number_sections: true
    keep_tex: true
    includes:
      in_header: "preamble.tex"
fontsize: "12pt"
indent: true
abstract: This paper details 6thsense, a mission to send a weather balloon to to edge of space in order to study Earth's atmosphere. The weather balloon is outfitted with a payload that includes a camera, inside and outside temperature sensors, a barometer, multiple GPS tracking devices, as well as a host of individual experiments. The balloon most likely reached an altitude of 20,000 meters, into the stratosphere and ozone layer. The balloon was successfully recovered after approximately one and a half hours of flight.
---

\newpage
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}
\tableofcontents
\newpage
\onehalfspacing

```{r setup, include=FALSE}
library(knitr)
library(mosaic)
library(gridExtra)
opts_chunk$set(echo = F)

```
#Introduction

There are four main layers to the atmosphere: the troposphere, stratosphere, mesosphere, and thermosphere. The **troposphere** is the section of the atmosphere from roughly 0km to 15km, and is the layer in which we live. It includes 75% of the mass of all gases in the atmosphere, consisting of 78% nitrogen gas, 21% oxygen, and less than 1% of argon, carbon dioxide, and other constituents. Because it is both the lowest in altitude and most massive layer, the pressure in the troposphere is the greatest: 100kPa. As you increase altitude in the troposphere, temperature decreases to about -60C. The **stratosphere** is the section from roughly 15km to 50km. The ozone layer, or the region in the atmosphere which absorbs most UV rays, is within the stratosphere at about 20km in altitude. The pressure in the stratosphere is 10kPa. The **mesosphere** is the region from 50km to 80km. The pressure in the mesosphere is 0.1 kPa. Finally, the **thermosphere** is the highest in altitude of the four layers, spanning from 80km into outer space. Here, the pressure is only 0.01kPa, as it is the least dense of the layers. The few air molecules in the thermosphere are heated to extremely high temperatures by the Sun’s radiation. However, because there are so few molecules, we cannot measure an appreciable difference in overall temperature, so the effective temperature of the thermosphere is less than that of the mesosphere.

The layers relevant to our experiment are the troposphere and stratosphere, as our balloon launch was designed to reach a maximum altitude of 30km (about the middle of the stratosphere but past the ozone layer). Therefore, we had to prepare the payload for temperatures as low as -60C and pressure as low as 10kPa (in the stratosphere). We expect to encounter the lowest temperature in the troposphere, around an altitude of 10km. As the balloon rises from sea level to the top of the troposphere, we expect to see temperature decrease because the ground absorbs most of the sun’s radiation (as opposed to the gas particles). However, as the balloon enters the stratosphere, we expect an increase in temperature due to the ozone layer’s blockage of UV rays. The UV rays that the ozone layer absorbs causes the gas particles in the stratosphere to heat up, so temperature increases with altitude. As the balloon gains altitude, the density of air around it decreases, so the pressure decreases.

The presence of greenhouse gasses in the atmosphere contribute to the warming of the earth’s surface and the troposphere because of the greenhouse effect. The greenhouse effect describes the trapping of heat within the earth’s atmosphere via the infrared absorption of certain molecules.

When infrared radiation enters the atmosphere, short-wave infrared radiation (700 -  1000 nm) does not effectively interact with greenhouse gasses. The photons instead pass through the atmosphere and reach the surface of the earth. Here, they are absorbed by the surface of the earth and get reemitted as a lower energy, longer-waved infrared photon. Greenhouse gasses can now absorb these lower energy photons and reemit them. This absorption and reemission warms the troposphere by increasing the kinetic energy the atmosphere. Additionally, greenhouse gasses can reemit photons back towards the surface of the earth, increasing the temperature of the earth’s surface. By impeding the tendency for long-wave radiation to leave the atmosphere, the greenhouse effect increase the temperature of the troposphere and surface.

The spectroscopy that explains the greenhouse effect concerns molecular vibration. Infrared photons excite vibrations in molecules. Since the vibrational frequency of atoms are quantized, only certain energies will excite vibration in a molecule. Different molecules can exhibit different vibrational modes. A molecule with more atoms has more vibrational modes. The infrared selection rule states that a vibrational mode is active when it is changing the dipole moment of the molecule. An active vibrational mode is a molecular vibration that can be excited by an infrared photon. Thus, infrared photons will only be absorbed when that molecule’s dipole moment is changing. Molecules such as nitrogen and oxygen will never exhibit a change in their dipole moment due to vibration because their only vibrational mode involves the singular oscillation of the distance between two atoms. Thus, they cannot absorb infrared radiation in order to partake in the greenhouse effect and are not considered greenhouse gasses. R On the other hand, many atmospheric molecules that have more than two atoms have more complex vibrational modes. These active vibrational mode gasses, such as nitrogen dioxide and water, are called greenhouse gasses.

As the greenhouse effect warms the planet, many factors accelerate the temperature rise by creating positive feedback loops. Positive feedback loops amplify any global increase in temperature. One positive feedback loops is created when rising temperatures melt glaciers. These white and reflective icy surfaces melt into dark oceans which absorb more radiation and heat up faster. This explains the approaching transition from a permanent to a seasonal ice cover in the Arctic Ocean. Thus temperature increases accelerate.^[https://www.nature.com/articles/s41598-017-08467-z] Another positive feedback loops involves the melting of permafrost. A quarter of the northern hemisphere is covered in permafrost that holds concentrated methane and carbon dioxide. Slight growths in temperature are enough to thaw the permafrost. When it melts, these greenhouse gasses are released into the atmosphere. Thus, the greenhouse effect amplifies and global temperature growth accelerates. ^[http://science.sciencemag.org/content/340/6129/183] Finally, humidity is also a positive feedback loop. As global temperatures rise, the atmosphere can hold more water vapor because less of it gets condensed. Water vapor is a greenhouse gas itself. Thus any increase in global temperatures will be amplified due to the water vapor positive feedback loop.

@@

TGFs are short, powerful bursts of gamma rays detected in the atmosphere. Each burst contains 10^17 to 10^19 gamma rays and last about 1 ms. They were first detected in the 1950s. The United States, concerned with the Soviet Union’s nuclear progress used satellites to detect gamma rays and track the enemy’s progress. After observing long, 5-10 second gamma ray glows (suspected to be Gamma ray Bursts (GRB) from stars), the U.S. sent more satellites to explore GRB further. In addition to detecting GRB, the data from these satellites led to the discovery of TGFs. When short, gamma ray bursts were analyzed for their source locations, researchers noticed that these burst mainly came from coastal sites on the equator. After recognizing a correlation with lightning prone areas and these short gamma ray bursts, C. T. R. Wilson’s theory concerning a Relativistic Runaway Electron Avalanche (RREA) was applied to discover TGFs.

Wilson’s RREA theory relies on a quantum properties of electrons. As the velocity of normal objects traveling through the atmosphere increases, the air resistance increases also. Therefore, in a constant force field, the object will reach a terminal velocity. For electrons however, while the air resistance initially increase with its speed, at high velocities, the air resistance on the electron begins to decrease. This is because at relativistic speeds, the electron can avoid interactions with other molecules. Thus, after the electron reaches a certain speed, it will never reach a terminal velocity and accelerate indefinitely as long as it is still in a force field. Electrons that escape terminal velocity are called runaway electrons. The second part of Wilson theory involves an explanation of the “Avalanche.” Runaway electrons can knock other electrons while accelerating and cause them to become runaway electrons also. This creates an exponential increase of runaway electrons as more and more electrons interact.

@@

**INSERT PRESSURE MODEL ANALYSIS HERE**

@@

# Experiments Conducted

@@

## Group Experiments
```{r experiments_conducted, echo=FALSE}
group_experiments <- data.frame("Sensor" = c("Real Time Clock", "GPS", "Barometer", "Temperature", "Humidity", "Temperature"), "Function" = c("Accurate Timing", "Location Data", "Altitude Data", "Outside Temperature", "Outside Humidity", "Inside Temperature"), "Model" = c("Sparkfun DS1307 RTC", "Sparkfun Venus GPS", "Vernier Gas Pressure Sensor", "Adafruit BME280", "Adafruit BME280", "Thermistor"))
kable(group_experiments)
```

@@

## Individual Experiments Conducted
```{r inidivduals, echo=F}
individuals <- data.frame("Team Member" = c("Alec Vercruysse", "Luke Bowsher", "Simon Oros", "Vivian Liu", "Caroline Pang", "John Kim"), "Experiment" = c("Spectrometer", "Accelerometer", "~~Geiger Counter~~*", "Light Intensity Sensors", "UVB Sensor", "Methane Sensor"))
kable(individuals)
```

*shortly before launch it was discovered that the Geiger Counter was non-operable and therefore it was taken out of the payload. Instead, Simon Oros focused on ____.

**INCLUDE INDIVIDUAL MOTIVATIONS HERE**

# Physical Design

@@

# Cut Down Mechanism

@@

# Electrical and Software Design

@@

# Tracking The Payload

@@

## The Global Positioning System

The Global Positioning System (GPS) is a system of satellites in Medium Earth Orbit around Earth which provide accurate location details to any device connecting to the network. The orbits of these 27 satellites, while not geostationary, ensure that four satellites are visible from any location on Earth at any given moment.^[http://www.astronomy.ohio-state.edu/~pogge/Ast162/Unit5/gps.html#note01] By evaluating the time taken for GPS signals to be sent down to earth in the form of radio waves traveling at the speed of light, a GPS receiver is able to calculate the distance between it and the satellite. Using this distance information gained from at least three satellites, a GPS receiver is then able to “trilaterate” to find its location relative to the satellites.^[http://www.physics.org/article-questions.asp?id=55]
 Given the distance between a satellite, a GPS unit must be located on a sphere extending outward with a radius of that distance. With information from three satellites, three spheres are given, and generally the intersection of three spheres results in two points, once which can be discarded due to the fact that it would not be physically possible for the GPS receiver to be located there.

The Department of Commerce has placed regulations on exports that could potentially threaten the United States. This includes GPS receivers that can provide data when they are above 18,000 meters or their speed is above 1000 knots, due to their potential use in Intercontinental Ballistic Missiles.^[http://ravtrack.com/GPStracking/cocom-gps-tracking-limits/469/]Some manufacturers have implemented these limits by shutting down the system when either condition is met, and some manufacturers have implemented these limits by shutting down the system only if both conditions are met. This is an important distinction, since high altitude balloons often exceed the height limit of 18,000 meters. Sparkfun’s Venus GPS module, which uses the Venus634FLPx GPS chipset, is known to allow altitudes of over 18,000 meters, given that the speed of the GPS receiver is not calculated to be more than 1000 knots.^[https://ukhas.org.uk/guides:gps_modules]

## The PicoAPRS Transceiver

The primary tracking system onboard the balloon was a PicoAPRS, developed by Taner Schenker (DB1NTO). This is a small transceiver that can receive and broadcast on the Automatic Packet Reporting System (APRS). Most importantly, it contains its own separate GPS chip that allows it to broadcast its location to the APRS network. This device is completely self contained, including its own 850MAh Lithium Ion battery, making it a perfect candidate for tracking -- even if all other systems on the balloon have failed, the PicoAPRS should still be able to transmit its location. Furthermore, the PicoAPRS provides a simple menu for device configuration, to enable the performance desired. Specifically, the PicoAPRS transmitted GPS beacons at a power of 1 Watt over intervals of 60 seconds. The PicoAPRS did this with Menlo School’s callsign, N6MLO, and a unique SSID, 9, that differentiated the packets sent by 6thsense’s PicoAPRS from other concurrent Menlo School balloon launches.

## The APRS Network and MIC-Encoding

A very important benefit of using the PicoAPRS is that it uses the Automatic Packet Reporting System (APRS). This amateur radio network, developed by Bob Bruninga (WB4APR) uses digipeaters to forward properly formatted broadcasted beacon packets to other stations in the area to effectively extend the range of a single transmitter. Rather than focusing on ensuring that a packet is received by all digipeaters, this protocol focuses on redundancy through packet “multiplication” by having each digipeater that receives the signal propagate it outward through the network. While the exact details of how many “hops” a packet can take through the network before transmission ends are dependent on the settings present in the header of the packet. This system became popular, especially for transmitting GPS data and tracking vehicles, and eventually specifications were developed to transfer packet data to the web through APRS-Internet Service (APRS-IS).^[http://www.aprs-is.net/] The aprs.fi website, which interfaces with APRS-IS and displays live telemetry data superimposed on google maps, provided a simple method of tracking the balloon during the launch.

The APRS network transmits packets through a single AX.25 data link protocol, which specifies the structure of frames to be sent through the network. Specifically, it uses Unnumbered Information (UI) frames, a type of AX.25 frame that is transmitted without any expectation of a response confirming reception, and reception is not guaranteed. APRS supports different encoding of these AX.25 UI-frames, however, to allow for different uses of the network. The PicoAPRS transmitter uses “Mic-Encoding” (MIC-E), a method that allows for extremely compressed packets that still contain position, course, speed, a message, and all relevant path settings and necessary headers. This achieved by writing compressed latitude information in the destination address field of the AX.25 frame, and compressed Longitude information in the frame’s Information Field. By transmitting extremely short packets that are still supported by the APRS network and aprs.fi, Mic-E can improve the reliability of the beacons transmitted.

@@

## The 900MHz Venus GPS Transciever

The backup tracking system involved sending GPS along with other telemetry data from the main arduino mega to the base station using Digikey 9XTend 900MHz transceivers.^[https://www.sparkfun.com/datasheets/Wireless/Zigbee/xtend-productmanual.pdf] The use of these transceivers provided a high power 1 Watt signal in a relatively clear band that let a novel communications protocol be implemented that allowed for ease of parsing. The 9XTend module interfaced easily with both balloon and base station arduinos through Universal Asynchronous Receiver-Transmitter (UART) Serial. Since the Arduino Mega provides three usable Serial ports, one was used to send data to the 9XTend Module. By taking advantage of Arduino’s Serial library, a library was developed that allowed the 9XTend modules to transmit ASCII encoded bytes, which, given some stream editing by the base station to remove ASCII control characters injected by the base 9XTend Module, is parsable as an ASCII byte stream by software on the base station.

### A Novel Transmission Protocol

@@

`******<value-1>------<value-2>------...------<value-n>`

Where `<value-i>` is an ASCII encoded, calibrated and human readable sensor value. This allows for simplex operation in which a base station can tune in at any time--if the base station loses reception or power for some reason, it can listen in for the next packet when it comes back online by simply waiting for the 6 stars indicating the start of a packet. Furthermore, data was encoded in ASCII bytes to let the output be human readable with little computation. This way, the output stream could be easily analyzed without having to convert bytes containing raw numerical values into ASCII text. This gets rid of the need to make specifications for negative numbers and floats, and the need to specify the size of transmitted data. By encoding each digit as an ASCII byte, it was simple to account for negative numbers by preceding the value with a dash just as is normally done with displayed numbers, and to account for floats by adding a decimal place behind the ones digit. Furthermore, more complicated sensor readings like GPS and Spectrometer output that contained ASCII formatted text could be sent as-is with zero need for manipulation. Since 12 values were transmitted (refer to Section 5) with uncompressed ASCII bytes, packet sizes were long: up to 240 bytes. Because of this, there were often small failures in the packet were a byte or series of bytes was not received. Such packets were still logged with the notion that it was still possible to extract most of the data out of them instead of having them completely go to waste. To ensure that the start of a packet could be recognized and separate values could be delineated properly, however, a full six bytes was used to mark the beginning of a transmission, and breaks separating values. This data was only transmitted about once every 12 seconds, partly to serve as flow control in order to not overload the transmitter’s buffer, and partly to conserve battery power, as the 9XTend module consumed 1 Watt while transmitting and was connected to the same battery as the Arduino Mega running all the sensors and SD logger.

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Sending a Balloon to the Edge of Space},
            pdfauthor={Luker Bowsher, John Kim, Vivian Liu, Simon Oros, Caroline Pang, Alec Vercruysse},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\usepackage[normalem]{ulem}
% avoid problems with \sout in headers with hyperref:
\pdfstringdefDisableCommands{\renewcommand{\sout}{}}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Sending a Balloon to the Edge of Space}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Luker Bowsher, John Kim, Vivian Liu, Simon Oros, Caroline Pang, Alec
Vercruysse}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{November 13, 2018}

\usepackage{setspace}
\usepackage{gensymb}

\begin{document}
\maketitle
\begin{abstract}
This paper details 6thsense, a mission to send a weather balloon to to
edge of space in order to study Earth's atmosphere. The weather balloon
is outfitted with a payload that includes a camera, inside and outside
temperature sensors, a barometer, multiple GPS tracking devices, as well
as a host of individual experiments. The balloon most likely reached an
altitude of 20,000 meters, into the stratosphere and ozone layer. The
balloon was successfully recovered after approximately one and a half
hours of flight.
\end{abstract}

\newpage

\setcounter{secnumdepth}{2} \setcounter{tocdepth}{2} \tableofcontents
\newpage
\onehalfspacing

\section{Introduction}\label{introduction}

There are four main layers to the atmosphere: the troposphere,
stratosphere, mesosphere, and thermosphere. The \textbf{troposphere} is
the section of the atmosphere from roughly 0km to 15km, and is the layer
in which we live. It includes 75\% of the mass of all gases in the
atmosphere, consisting of 78\% nitrogen gas, 21\% oxygen, and less than
1\% of argon, carbon dioxide, and other constituents. Because it is both
the lowest in altitude and most massive layer, the pressure in the
troposphere is the greatest: 100kPa. As you increase altitude in the
troposphere, temperature decreases to about -60C. The
\textbf{stratosphere} is the section from roughly 15km to 50km. The
ozone layer, or the region in the atmosphere which absorbs most UV rays,
is within the stratosphere at about 20km in altitude. The pressure in
the stratosphere is 10kPa. The \textbf{mesosphere} is the region from
50km to 80km. The pressure in the mesosphere is 0.1 kPa. Finally, the
\textbf{thermosphere} is the highest in altitude of the four layers,
spanning from 80km into outer space. Here, the pressure is only 0.01kPa,
as it is the least dense of the layers. The few air molecules in the
thermosphere are heated to extremely high temperatures by the Sun's
radiation. However, because there are so few molecules, we cannot
measure an appreciable difference in overall temperature, so the
effective temperature of the thermosphere is less than that of the
mesosphere.

The layers relevant to our experiment are the troposphere and
stratosphere, as our balloon launch was designed to reach a maximum
altitude of 30km (about the middle of the stratosphere but past the
ozone layer). Therefore, we had to prepare the payload for temperatures
as low as -60C and pressure as low as 10kPa (in the stratosphere). We
expect to encounter the lowest temperature in the troposphere, around an
altitude of 10km. As the balloon rises from sea level to the top of the
troposphere, we expect to see temperature decrease because the ground
absorbs most of the sun's radiation (as opposed to the gas particles).
However, as the balloon enters the stratosphere, we expect an increase
in temperature due to the ozone layer's blockage of UV rays. The UV rays
that the ozone layer absorbs causes the gas particles in the
stratosphere to heat up, so temperature increases with altitude. As the
balloon gains altitude, the density of air around it decreases, so the
pressure decreases.

The presence of greenhouse gasses in the atmosphere contribute to the
warming of the earth's surface and the troposphere because of the
greenhouse effect. The greenhouse effect describes the trapping of heat
within the earth's atmosphere via the infrared absorption of certain
molecules.

When infrared radiation enters the atmosphere, short-wave infrared
radiation (700 - 1000 nm) does not effectively interact with greenhouse
gasses. The photons instead pass through the atmosphere and reach the
surface of the earth. Here, they are absorbed by the surface of the
earth and get reemitted as a lower energy, longer-waved infrared photon.
Greenhouse gasses can now absorb these lower energy photons and reemit
them. This absorption and reemission warms the troposphere by increasing
the kinetic energy the atmosphere. Additionally, greenhouse gasses can
reemit photons back towards the surface of the earth, increasing the
temperature of the earth's surface. By impeding the tendency for
long-wave radiation to leave the atmosphere, the greenhouse effect
increase the temperature of the troposphere and surface.

The spectroscopy that explains the greenhouse effect concerns molecular
vibration. Infrared photons excite vibrations in molecules. Since the
vibrational frequency of atoms are quantized, only certain energies will
excite vibration in a molecule. Different molecules can exhibit
different vibrational modes. A molecule with more atoms has more
vibrational modes. The infrared selection rule states that a vibrational
mode is active when it is changing the dipole moment of the molecule. An
active vibrational mode is a molecular vibration that can be excited by
an infrared photon. Thus, infrared photons will only be absorbed when
that molecule's dipole moment is changing. Molecules such as nitrogen
and oxygen will never exhibit a change in their dipole moment due to
vibration because their only vibrational mode involves the singular
oscillation of the distance between two atoms. Thus, they cannot absorb
infrared radiation in order to partake in the greenhouse effect and are
not considered greenhouse gasses. R On the other hand, many atmospheric
molecules that have more than two atoms have more complex vibrational
modes. These active vibrational mode gasses, such as nitrogen dioxide
and water, are called greenhouse gasses.

As the greenhouse effect warms the planet, many factors accelerate the
temperature rise by creating positive feedback loops. Positive feedback
loops amplify any global increase in temperature. One positive feedback
loops is created when rising temperatures melt glaciers. These white and
reflective icy surfaces melt into dark oceans which absorb more
radiation and heat up faster. This explains the approaching transition
from a permanent to a seasonal ice cover in the Arctic Ocean. Thus
temperature increases accelerate.\footnote{\url{https://www.nature.com/articles/s41598-017-08467-z}}
Another positive feedback loops involves the melting of permafrost. A
quarter of the northern hemisphere is covered in permafrost that holds
concentrated methane and carbon dioxide. Slight growths in temperature
are enough to thaw the permafrost. When it melts, these greenhouse
gasses are released into the atmosphere. Thus, the greenhouse effect
amplifies and global temperature growth accelerates. \footnote{\url{http://science.sciencemag.org/content/340/6129/183}}
Finally, humidity is also a positive feedback loop. As global
temperatures rise, the atmosphere can hold more water vapor because less
of it gets condensed. Water vapor is a greenhouse gas itself. Thus any
increase in global temperatures will be amplified due to the water vapor
positive feedback loop.

@@

TGFs are short, powerful bursts of gamma rays detected in the
atmosphere. Each burst contains 10\^{}17 to 10\^{}19 gamma rays and last
about 1 ms. They were first detected in the 1950s. The United States,
concerned with the Soviet Union's nuclear progress used satellites to
detect gamma rays and track the enemy's progress. After observing long,
5-10 second gamma ray glows (suspected to be Gamma ray Bursts (GRB) from
stars), the U.S. sent more satellites to explore GRB further. In
addition to detecting GRB, the data from these satellites led to the
discovery of TGFs. When short, gamma ray bursts were analyzed for their
source locations, researchers noticed that these burst mainly came from
coastal sites on the equator. After recognizing a correlation with
lightning prone areas and these short gamma ray bursts, C. T. R.
Wilson's theory concerning a Relativistic Runaway Electron Avalanche
(RREA) was applied to discover TGFs.

Wilson's RREA theory relies on a quantum properties of electrons. As the
velocity of normal objects traveling through the atmosphere increases,
the air resistance increases also. Therefore, in a constant force field,
the object will reach a terminal velocity. For electrons however, while
the air resistance initially increase with its speed, at high
velocities, the air resistance on the electron begins to decrease. This
is because at relativistic speeds, the electron can avoid interactions
with other molecules. Thus, after the electron reaches a certain speed,
it will never reach a terminal velocity and accelerate indefinitely as
long as it is still in a force field. Electrons that escape terminal
velocity are called runaway electrons. The second part of Wilson theory
involves an explanation of the ``Avalanche.'' Runaway electrons can
knock other electrons while accelerating and cause them to become
runaway electrons also. This creates an exponential increase of runaway
electrons as more and more electrons interact.

@@

\textbf{INSERT PRESSURE MODEL ANALYSIS HERE}

@@

\section{Experiments Conducted}\label{experiments-conducted}

@@

\subsection{Group Experiments}\label{group-experiments}

\begin{longtable}[]{@{}lll@{}}
\toprule
Sensor & Function & Model\tabularnewline
\midrule
\endhead
Real Time Clock & Accurate Timing & Sparkfun DS1307 RTC\tabularnewline
GPS & Location Data & Sparkfun Venus GPS\tabularnewline
Barometer & Altitude Data & Vernier Gas Pressure Sensor\tabularnewline
Temperature & Outside Temperature & Adafruit BME280\tabularnewline
Humidity & Outside Humidity & Adafruit BME280\tabularnewline
Temperature & Inside Temperature & Thermistor\tabularnewline
\bottomrule
\end{longtable}

@@

\subsection{Individual Experiments
Conducted}\label{individual-experiments-conducted}

\begin{longtable}[]{@{}ll@{}}
\toprule
Team.Member & Experiment\tabularnewline
\midrule
\endhead
Alec Vercruysse & Spectrometer\tabularnewline
Luke Bowsher & Accelerometer\tabularnewline
Simon Oros & \sout{Geiger Counter}*\tabularnewline
Vivian Liu & Light Intensity Sensors\tabularnewline
Caroline Pang & UVB Sensor\tabularnewline
John Kim & Methane Sensor\tabularnewline
\bottomrule
\end{longtable}

*shortly before launch it was discovered that the Geiger Counter was
non-operable and therefore it was taken out of the payload. Instead,
Simon Oros focused on \_\_\_\_.

\textbf{INCLUDE INDIVIDUAL MOTIVATIONS HERE}

\section{Physical Design}\label{physical-design}

@@

\section{Cut Down Mechanism}\label{cut-down-mechanism}

@@

\section{Electrical and Software
Design}\label{electrical-and-software-design}

@@

\section{Tracking The Payload}\label{tracking-the-payload}

@@

\subsection{The Global Positioning
System}\label{the-global-positioning-system}

The Global Positioning System (GPS) is a system of satellites in Medium
Earth Orbit around Earth which provide accurate location details to any
device connecting to the network. The orbits of these 27 satellites,
while not geostationary, ensure that four satellites are visible from
any location on Earth at any given moment.\footnote{\url{http://www.astronomy.ohio-state.edu/~pogge/Ast162/Unit5/gps.html\#note01}}
By evaluating the time taken for GPS signals to be sent down to earth in
the form of radio waves traveling at the speed of light, a GPS receiver
is able to calculate the distance between it and the satellite. Using
this distance information gained from at least three satellites, a GPS
receiver is then able to ``trilaterate'' to find its location relative
to the satellites.\footnote{\url{http://www.physics.org/article-questions.asp?id=55}}
Given the distance between a satellite, a GPS unit must be located on a
sphere extending outward with a radius of that distance. With
information from three satellites, three spheres are given, and
generally the intersection of three spheres results in two points, once
which can be discarded due to the fact that it would not be physically
possible for the GPS receiver to be located there.

The Department of Commerce has placed regulations on exports that could
potentially threaten the United States. This includes GPS receivers that
can provide data when they are above 18,000 meters or their speed is
above 1000 knots, due to their potential use in Intercontinental
Ballistic Missiles.\footnote{\url{http://ravtrack.com/GPStracking/cocom-gps-tracking-limits/469/}}Some
manufacturers have implemented these limits by shutting down the system
when either condition is met, and some manufacturers have implemented
these limits by shutting down the system only if both conditions are
met. This is an important distinction, since high altitude balloons
often exceed the height limit of 18,000 meters. Sparkfun's Venus GPS
module, which uses the Venus634FLPx GPS chipset, is known to allow
altitudes of over 18,000 meters, given that the speed of the GPS
receiver is not calculated to be more than 1000 knots.\footnote{\url{https://ukhas.org.uk/guides:gps_modules}}

\subsection{The PicoAPRS Transceiver}\label{the-picoaprs-transceiver}

The primary tracking system onboard the balloon was a PicoAPRS,
developed by Taner Schenker (DB1NTO). This is a small transceiver that
can receive and broadcast on the Automatic Packet Reporting System
(APRS). Most importantly, it contains its own separate GPS chip that
allows it to broadcast its location to the APRS network. This device is
completely self contained, including its own 850MAh Lithium Ion battery,
making it a perfect candidate for tracking -- even if all other systems
on the balloon have failed, the PicoAPRS should still be able to
transmit its location. Furthermore, the PicoAPRS provides a simple menu
for device configuration, to enable the performance desired.
Specifically, the PicoAPRS transmitted GPS beacons at a power of 1 Watt
over intervals of 60 seconds. The PicoAPRS did this with Menlo School's
callsign, N6MLO, and a unique SSID, 9, that differentiated the packets
sent by 6thsense's PicoAPRS from other concurrent Menlo School balloon
launches.

\subsection{The APRS Network and
MIC-Encoding}\label{the-aprs-network-and-mic-encoding}

A very important benefit of using the PicoAPRS is that it uses the
Automatic Packet Reporting System (APRS). This amateur radio network,
developed by Bob Bruninga (WB4APR) uses digipeaters to forward properly
formatted broadcasted beacon packets to other stations in the area to
effectively extend the range of a single transmitter. Rather than
focusing on ensuring that a packet is received by all digipeaters, this
protocol focuses on redundancy through packet ``multiplication'' by
having each digipeater that receives the signal propagate it outward
through the network. While the exact details of how many ``hops'' a
packet can take through the network before transmission ends are
dependent on the settings present in the header of the packet. This
system became popular, especially for transmitting GPS data and tracking
vehicles, and eventually specifications were developed to transfer
packet data to the web through APRS-Internet Service
(APRS-IS).\footnote{\url{http://www.aprs-is.net/}} The aprs.fi website,
which interfaces with APRS-IS and displays live telemetry data
superimposed on google maps, provided a simple method of tracking the
balloon during the launch.

The APRS network transmits packets through a single AX.25 data link
protocol, which specifies the structure of frames to be sent through the
network. Specifically, it uses Unnumbered Information (UI) frames, a
type of AX.25 frame that is transmitted without any expectation of a
response confirming reception, and reception is not guaranteed. APRS
supports different encoding of these AX.25 UI-frames, however, to allow
for different uses of the network. The PicoAPRS transmitter uses
``Mic-Encoding'' (MIC-E), a method that allows for extremely compressed
packets that still contain position, course, speed, a message, and all
relevant path settings and necessary headers. This achieved by writing
compressed latitude information in the destination address field of the
AX.25 frame, and compressed Longitude information in the frame's
Information Field. By transmitting extremely short packets that are
still supported by the APRS network and aprs.fi, Mic-E can improve the
reliability of the beacons transmitted.

@@

\subsection{The 900MHz Venus GPS
Transciever}\label{the-900mhz-venus-gps-transciever}

The backup tracking system involved sending GPS along with other
telemetry data from the main arduino mega to the base station using
Digikey 9XTend 900MHz transceivers.\footnote{\url{https://www.sparkfun.com/datasheets/Wireless/Zigbee/xtend-productmanual.pdf}}
The use of these transceivers provided a high power 1 Watt signal in a
relatively clear band that let a novel communications protocol be
implemented that allowed for ease of parsing. The 9XTend module
interfaced easily with both balloon and base station arduinos through
Universal Asynchronous Receiver-Transmitter (UART) Serial. Since the
Arduino Mega provides three usable Serial ports, one was used to send
data to the 9XTend Module. By taking advantage of Arduino's Serial
library, a library was developed that allowed the 9XTend modules to
transmit ASCII encoded bytes, which, given some stream editing by the
base station to remove ASCII control characters injected by the base
9XTend Module, is parsable as an ASCII byte stream by software on the
base station.

\subsubsection{A Novel Transmission
Protocol}\label{a-novel-transmission-protocol}

@@

\texttt{******\textless{}value-1\textgreater{}-\/-\/-\/-\/-\/-\textless{}value-2\textgreater{}-\/-\/-\/-\/-\/-...-\/-\/-\/-\/-\/-\textless{}value-n\textgreater{}}
